Epistemic Semantics in Guarded String Models

### 2020-09-26 tex/sdraft.tex
Tiny edits in sdraft.

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a
natural language.  A proposition is a set or class of {\em possible
  worlds}.  These are often assumed to be things with the same nature
and complexity as the world we occupy \citep{lewis1986plurality}.  But
alternatively, one can work with small idealized models, in order to
illustrate and test ideas.  The point of this paper is to scale up toy
or idealized models to countable sets of worlds, and to constructive
and computable modeling of epistemic alternatives for agents.  We
describe a certain systematic way of defining such models, and
illustrate how to apply them in natural language semantics.  The focus
is on epistemic semantics and clausal embedding. The fundamental move
is to identify possible worlds with strings of primitive events, so
that propositions are sets of strings. An advantage in this is that it
allows for a mathematical description of an algebra of propositions,
coupled with a computational representation using either lazy lists of
strings, or finite state machines that describe sets of strings.

@@

\ehc{I'm not sure if these sentences are the most interesting anymore!}

@@

Formulas like the ones in \ex{event1} that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between atoms, as defined in Figure
\ref{effectfigure}.  The atoms they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Amy peeking at heads) be as
specified in \ex{effect}. Then ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\varphi}$
and the relation on atoms for the event $a_1$ are as given at
the bottom in \ex{effect}.

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, and events of the agents
semi-privately looking at the coin. The line beginning with {\tt
  state} enumerates ${\sf{B}}$. The line beginning with {\tt restrict}
gives the state formula.  The lines beginning with {\tt event} declare
events and their effect formulas.  Finally the lines beginning with
{\tt agent} define {\em event alternative} relations for agents.  Each
clause with an arrow has a single event symbol on the left, and a
disjunction of alternative events on the right of the arrow. The
interpretation of Amy's alternatives for $b_1$ (Bob peeks at heads),
is that when $b_1$ happens, for Amy either $b_1$ or $b_0$ (Bob peeks
at tails) could be happening. Her alternatives for $a_{01}$ and
$a_{10}$ (she turns the coin over) are $a_{10}$ and $a_{01}$,
indicating that she doesn't know, \emph{a priori}, whether she's
turning the coin from $H$ to $T$ or from $T$ to $H$. Similarly, Bob
secretly turns the coin over, in $b_{10}$ or $b_{01}$, she doesn't
know anything has happened, so her alternatives are the no-operation
events $o_1$ and $o_0$ for heads-worlds and tails-worlds
respectively. Bob's alternatives are the same, \emph{mutatis mutandi}.

@@

This paper focuses on defining a concrete possible worlds model from
an Epik specification.  The models are an extension of guarded-string
models for Kleene Algebra with Tests (KAT). This is an algebraic
theory that has model classes including guarded string models,
relational models, finite models, and matrix models.  Our definitions
and notation follow \cite{kozen2001automata}.  We add syntax and
semantics is included to cover multi-agent epistemic semantics.

@@

In the coin example, as we already saw in \ex{effect},
$\mathcal{A}^\varphi_{\mathsf B}$ is $\{h\bar{t},\bar{h}t\}$, for which we use
the shorthand $\{H,T\}$.  A guarded string over ${\sf{P}}$ and ${\sf{B}}$ is a
strings of events from ${\sf{P}}$, alternating with atoms over ${\sf{B}}$, and
beginning and ending with atoms.  \ex{gs1} gives the encoding as guarded strings
of the worlds in \ex{elevator1} and \ex{coin1}.  The length of a guarded string
$p$, written $|p|$ is the number of events in $p$.  An atom such as $H$ is a
guarded string of length $0$. Correspondingly $|w_1| = 0$, $|w_2| = 1$ and
$|w_3| = 2$.

@@

Rather than individual guarded strings, elements of a guarded string
model for KAT are sets of guarded strings.  In our application, these
elements have the interpretation of propositions, which are sets of possible
worlds.  In a free guarded string model for KAT, any event can be
adjacent to any atom in a guarded string that is an element of the
underlying set for the algebra.  We instead impose the constraints
coming from the state and effect formulas.  \ex{wf1} defines the
well-formed guarded strings determined by and Epik specification.  Condition (i)
says that each atom is consistent with the state constraint, and condition (ii)
says that each constituent token event $\alpha_ie_i\alpha_{i+1}$ is
consistent with the effect constraint on $e_i$.\footnote{
An alternative is to define equations such as
$\bar{\phi}=0$ (from the state formula $\phi$)
and $a_1\,{=}\,ha_{1}h$ (from the effect formula $h\mathbin{:}h$ for event $a_1$),
and construct a quotient algebra from the equivalence relation generated
by these equations.  This results in equating sets of guarded strings in the free algebra that
differ by guarded strings that are ill-formed according to the state and
effect formulas.  In the development in the text, we instead use
a set of guarded strings that are well-formed according to the state
and effect formulas as the representative of the equivalence class.
}

@@

Well-formed guarded strings have the interpretation of worlds in the application
to natural-language semantics.  The set of possible worlds in the Kripke frame
determined by an Epik specification is the set of well-formed guarded strings.
At this point, we could say that any set of worlds is a proposition, so that the
set of propositions is the power set of the set of worlds
\citep{montague1975formal,gallin1975intensional}.  We will instead define a more
restrictive set of propositions corresponding to the regular sets of
strings. This is deferred to the next section.  Certain sets of well-formed
guarded strings have the addional interpretation of event types.  An event-type
is something that can ``happen'' is different worlds. For example, $a_1$ has the
event type $\{H\,a_1\,H\}$, and $a_0$ has the even type $\{T\,a_0\,T\}$. The
event types for $b_0$ and $b_1$ are analogous.

The construction so far defines a set of worlds from an Epik specification.  Normally
the set is countably infinite, though some choices of effect
formulas can result in a finite set of worlds.  The next step is to define an
alternative relation $R_a$ on worlds for each agent $a$. This will result in a Kripke
frame $\tuple{W,R_1,...,R_n}$ consisting of a set of worlds, and a world-alternative
relation for each agent \citep{kripke1963semantical}.  An Epik specification defines
an alternative relation on bare events for each agent $a$, which we notate as
$\hat{R}_{a}$.  This should be lifted to a relation ${R}_a$ on worlds.  The basic
idea is that when a world $w$ is incremented with an event $e$, in the resulting
world $w \diamond e$, epistemic alternatives for agent $a$ are of the form $w'
\diamond e'$, where $w'$ is an alternative to for $a$ in $w$, and $e'$ is and
event-alternative to $e$ for $a$.\footnote{In this it is important that the
  event-alternative relation for an agent is constant across worlds. We anticipate
  that the definition given here produces results equivalent to what is found in
  literature on event alternatives in dynamic epistemic semantics, though we have not
  verified this.  That literature primarily focuses on mapping an epistemic model for
  a single time and situation to another, and uses general first-order models, rather
  than guarded string models. See \citet{baltag1999logic}, \citet{van2007dynamic}, and
  articles in \citet{ditmarsch2015handbook}.  Previous literature is motivated by epistemic
  logic and AI planning, rather computable possible worlds models in natural language semantics.
} This needs to be
implemented in a way that takes account of pre- and post-conditions for events. For
this, our approach is to refer the definition of well-formed guarded strings.
\ex{raise1} defines a relation on worlds from a relation on bare events.

\begin{example}  \label{raise1}
  Let $W$ be a set of guarded strings over events P and primitive
  tests B, and $\hat{R}$ be a relation on $P$. The corresponding
  relation ${R}$ on $W$ holds between a guarded string
\(
\alpha_0e_0 ... e_n\alpha_{n+1} 
\)
in $W$ and a guarded string  $q$ iff $q$ is an element of $W$ and is of the form
\(
\alpha_0'e_0' ... e_n'\alpha_{n+1}', 
\), where for $0 \leq n$, $\element{\tuple{e_i,e'_i}}{\hat{R}}$.
\end{example}

@@

\section{The logical language of Epistemic KAT}
The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for KAT, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  } To this we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  Intuitively, $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1
  ... \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin ends heads-up\footnote{A mapping from English sentences to logical
  terms in epistemic KAT is presented in Section 6}.

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}
\ehc{I took the `that's out of these sentences. I remember from class
  that even if they are assigned null in the PF they still exist in
  the LF. Is it okay to leave it out here?}

\subsection{Standard Epistemic Modalities}

Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$ into
Epik\footnote{The choice to refer to $\Diamond_a$ as an ``epistemic''
  primitive rather than a ``doxastic'' primitive, is that it is
  essential in constructing both the doxastic and epistemic
  operators. A ``doxastic'' moniker would suggest the
  unrepresentability of epistemic modalities}. These encodings are
shown below:

@@

To construct different types of reasoners (e.g. accurate, inaccurate,
etc), users of Epik can construct an agents alternative relation
accordingly. Observe the discussion of modal axioms \textbf{T} and
\textbf{D} in Section~\ref{sec:axioms}.

\subsection{Guarded String Interpretation}
\newcommand{\epikspec}{\test,\event,\varphi,\zeta}
\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

A term $p$ of the logical language is interpreted as a set of guarded
strings $\sem{p}^{\epikspec}$, where superscript captures dependence
on an Epik specification.  Figure \ref{fig:interp2} defines the
interpretation.  The interpretation $\sem{ 1
}^{\epikspec}$ of the multiplicative identity 1 is the set of atoms that
satisfy the state constraint $\varphi$.  Where $b$ is a primitive
Boolean, $\sem{ b }^{\epikspec}$ is the set of atoms that satisfy the
state constraint and where $b$ is true.  Where $e$ is a primitive
event, $\sem{ e }^{\epikspec}$ is the set of guarded strings that have
the form of $e$ flanked by compatible atoms, as determined by the
event formula $\zeta_e$. The product $pq$ is
interpreted with fusion product raised to sets of guarded strings.
Kleene star is interpreted as the union of exponents ($p^n$ is
the $n$-times product of $p$ with itself, with $p^0 = 1$).
Propositional complement is complement relative to the set of worlds.
The epistemic formula $\Diamond_a p$ is interpreted with Kripke
semantics for epistemic modality, as the pre-image of the embedded
proposition under the world-alternative relation ${R}_a$.

@@

\subsection{Axiomatic Classification}
\label{sec:axioms}

To situate our logic as a modal logic, we investigate the soundness of
the standard modal axioms~\cite{????} in Epik's semantics. Some of
these standard axioms (\ex{valid-axioms}) hold all the time, i.e. they
are \emph{valid}, and the remaining axioms are valid when $\hat{R}_a$
has a certain shape. The axioms in \ex{conditionally-valid-axioms}
have a nontrivial condition validity condition.

@@

First, the space of worlds is a set of ordinary (as opposed to
guarded) strings.  Bit sequences (sequences of $0$'s and $1$'s) encode
atoms, as before, these alternate with event symbols to encode a
world. \ex{gs1fst} gives the encoding of worlds $w_2$ and $w_3$ from
the example. The valuation of $h$ preceeds the valuation of $t$, so
$1~0$ indicates the valuation $h\bar{t}$.

%% unnecessary detail
%% A string is a finite sequence of symbols, and {\tt 0}, {\tt 1}, {\tt
%%   u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
%% multi-character symbols that are found in implementations of the
%% finite state calculus. The multi-character symbol {\tt a0} is an
%% element of the alphabet that has no connection with the element {\tt
%%   a}.

@@

\begin{example}   \label{gs1fst}
  \renewcommand{\arraystretch}{1.2}
\begin{tabular}[t]{cl}
%% {\textit{World}} & {\textit{String}} \\
% $v_1$ & $ 0 \,u\,  1 \,u\,  2 \,d\,  1 \,u\,  2$ \\
%% $w_1$ & {\tt 1 0} \\
$w_2$ & {\tt 1 0 a1 1 0} \\
$w_3$ & {\tt 1 0 a1 1 0 b1 1 0} \\  
  \end{tabular}
    \renewcommand{\arraystretch}{1}
\end{example}

@@

Now, translating the Epik terms $0$, $1$, $b$, and $e$ are
straightforward: we simply convert the atoms as previously described,
ensuring to decorate the events $e$ with their compatible atoms
first. For example $a_1$ becomes $\{01a_101\}$ Fst also has built-in notions of union (\texttt{|}),
intersection (\texttt{\&}), and complement (\texttt{\~}). Translating
the the remaining operators requires more work.

Implementing fusion product using Fst's standard string concatenation
(denoted by juxatposition $\texttt{X} \texttt{Y}$) requires more
care. Naively concatenating strings with atoms (Boolean vectors) at
both ends doubles the atoms at the juncture, and does not enforce the
requisite atom equality.

To implement the fusion product, we define the binary product
operation \texttt{Cn}, which combines strings in the underlying string
algebra, remove strings with non-matching atoms, and then delete the
second of two tests to create well-formed guarded strings (see Figure
\ref{fstfigure1}). The set {\tt Wf0} describes pairs of equal atoms as
the complement of set of unequal atom pairs, implemented using Fst's
containment operator \texttt{\$}. Then. to delete the second of the
two (matching) pairs, the \texttt{Squash} operator uses Fst's rewrite
notation to delete atoms (\texttt{St}) that are preceded by another
atom. \footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used
  with relation composition and relation domain and co-domain.  They
  are restricted in that the complement and set difference for
  non-equal length relations is not defined.  The epistemic alterative
  relations that are defined in Figure -- are equal-length relations.
} The squash operator is applied via the relational composition
\texttt{.o.} and codomain \texttt{.l} operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

Kleene star is defined in a similar way (see Figure \ref{fstfigure1}),
unioning the set of atoms \texttt{St} with \texttt{Kpl} which uses
Fst's non-empty iteration operator \texttt{+} and normalizes the
result in the same way as \texttt{Cn}.

@@

\begin{hangingpar}
{\tt{define RelKst(R)
    [St\hspace{1pt}.x.\hspace{1pt}St]\hspace{1pt}|\hspace{1pt}Kpl(X);}}\\ The Fst
operation {\tt{.x.}} is Cartesian product.  {\tt{R.i}} is the inverse of relation
{\tt{R}}.
\end{hangingpar}
\end{tcolorbox}
\caption{Definition in Fst of the Kleene concatenation closure of
  a relation between guarded strings.}
\label{fstfigure2}
\end{figure}

@@

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for each
agent. This determines a relation in the guarded string algebra a relation between
bare events decorated with compatible tests.  For agent Amy, this is the relation
described in \ex{alyrel1} as a set of ordered pairs.

\vspace{-2mm}
\begin{example} \label{alyrel1}
 \hspace{-4mm} \(
 \left\{
 \begin{array}{l}
   \pair{10a_110}{10a_110},  \pair{01a_001}{01a_001}, \\
   \pair{10b_110}{10b_110},  \pair{10b_110}{01b_001}, \\
   \pair{01b_001}{10b_110},  \pair{01b_001}{01b_001} \\
 \end{array}
 \right\}
 \) 
\end{example}

% Fixes junky vertical spacing of the set above.
% Possibly remove it in the final version.
\raggedbottom

The relation on decorated events needs to be generalized to a relation
of worlds.  The principle for this is that an epistemic alternative to
a world of the form $we$ is a world of the form $vd$, where $v$ is a
world-alternative to $w$, $d$ is an event-alternative to $e$, and $vd$
is defined (i.e.  the world alternative $v$ satisfies the
pre-conditions of the event alternative $d$).  This principle is found
in earlier literature (Moore 198x, Baltag, Moss and Solecki 20xx).

>>>>

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a
natural language.  A proposition is a set or class of {\em possible
  worlds}.  These are often assumed to be things with the same nature
and complexity as the world we occupy \citep{lewis1986plurality}.  But
alternatively, one can work with small idealized models, in order to
illustrate and test ideas.  The point of this paper is to scale up toy
or idealized models to countable sets of worlds, and to constructive
and computable modeling of epistemic alternatives for agents.  We
describe a systematic way of defining such models, and
illustrate how to apply them in natural language semantics.  The focus
is on epistemic semantics and clausal embedding. The fundamental move
is to identify possible worlds with strings of primitive events, so
that propositions are sets of strings. An advantage in this is that it
allows for a mathematical description of an algebra of propositions,
coupled with a computational representation using either lazy lists of
strings, or finite state machines that describe sets of strings.

@@

%\ehc{I'm not sure if these sentences are the most interesting anymore!}

@@

Formulas that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between atoms, as defined in Figure
\ref{effectfigure}.  The atoms they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Amy peeking at heads) be as
specified in \ex{effect}. Then ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\varphi}$
and the relation on atoms for the event $a_1$ are as given at
the bottom in \ex{effect}.

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, events of the agents
semi-privately looking at the coin, and events of secretly turning the coin.
The line beginning with {\tt
  state} enumerates ${\sf{B}}$. The line beginning with {\tt restrict}
gives the state formula.  The lines beginning with {\tt event} declare
events and their effect formulas.  Finally the lines beginning with
{\tt agent} define {\em event alternative} relations for agents.  Each
clause with an arrow has a single event symbol on the left, and a
disjunction of alternative events on the right of the arrow. The
interpretation of Amy's alternatives for $b_1$ (Bob peeks at heads),
is that when $b_1$ happens, for Amy either $b_1$ or $b_0$ (Bob peeks
at tails) could be happening. Her alternatives for $a_{01}$ and
$a_{10}$ (she turns the coin over) are $a_{10}$ and $a_{01}$,
indicating that she doesn't know, \emph{a priori}, whether she's
turning the coin from $H$ to $T$ or from $T$ to $H$. Similarly, Bob
secretly turns the coin over, in $b_{10}$ or $b_{01}$, she doesn't
know anything has happened, so her alternatives are the no-operation
events $o_1$ and $o_0$ for heads-worlds and tails-worlds
respectively. Bob's alternatives are the same, \emph{mutatis mutandi}.

@@

This paper focuses on defining a concrete possible worlds model from
an Epik specification.  The models are an extension of guarded-string
models for Kleene Algebra with Tests (KAT). This is an algebraic
theory that has model classes including guarded string models,
relational models, finite models, and matrix models.  Our definitions
and notation follow \cite{kozen2001automata}.  We add syntax and
semantics to cover multi-agent epistemic semantics.

@@

In the coin example, as we already saw in \ex{effect},
$\mathcal{A}^\varphi_{\mathsf B}$ is $\{h\bar{t},\bar{h}t\}$, for
which we use the shorthand $\{H,T\}$.  A guarded string over
${\sf{P}}$ and ${\sf{B}}$ is a strings of events from ${\sf{P}}$,
alternating with atoms over ${\sf{B}}$, and beginning and ending with
atoms.  In this construction, $w_1=H$, $w_2=Ha_1H$, $w_3=Ha_1Hb_1H$,
and $w_4=Ha_1Hb_1Ha_{10}Tb_{01}H$.

@@

Rather than individual guarded strings, elements of a guarded string
model for KAT are sets of guarded strings.  In our application, these
elements have the interpretation of propositions, which are sets of possible
worlds.  In a free guarded string model for KAT, any event can be
adjacent to any atom in a guarded string that is an element of the
underlying set for the algebra.  We instead impose the constraints
coming from the state and effect formulas.  \ex{wf1} defines the
well-formed guarded strings determined by an Epik specification.  Condition (i)
says that each atom is consistent with the state constraint, and condition (ii)
says that each constituent token event $\alpha_ie_i\alpha_{i+1}$ is
consistent with the effect constraint on $e_i$.\footnote{
An alternative is to define equations such as
$\bar{\phi}=0$ (from the state formula $\phi$)
and $a_1\,{=}\,ha_{1}h$ (from the effect formula $h\mathbin{:}h$ for event $a_1$),
and construct a quotient algebra from the equivalence relation generated
by these equations.  This results in equating sets of guarded strings in the free algebra that
differ by guarded strings that are ill-formed according to the state and
effect formulas.  In the development in the text, we instead use
a set of guarded strings that are well-formed according to the state
and effect formulas as the representative of the equivalence class.
}

@@

Well-formed guarded strings have the interpretation of worlds in the application
to natural-language semantics.  The set of possible worlds in the Kripke frame
determined by an Epik specification is the set of well-formed guarded strings.
At this point, we could say that any set of worlds is a proposition, so that the
set of propositions is the power set of the set of worlds
\citep{montague1975formal,gallin1975intensional}.  We will instead define a more
restrictive set of propositions corresponding to the regular sets of
strings. This is deferred to the next section.  Certain sets of well-formed
guarded strings have the addional interpretation of event types.  An event-type
is something that can ``happen'' is different worlds. For example, $a_1$ has the
event type $\{H\,a_1\,H\}$, and $a_0$ has the event type $\{T\,a_0\,T\}$.

The construction so far defines a set of worlds from an Epik specification.  Normally
the set is countably infinite, though some choices of effect formulas can result in a
finite set of worlds.  The next step is to define an alternative relation $R_a$ on
worlds for each agent $a$. This will result in a Kripke frame $\tuple{W,R_1,...,R_n}$
consisting of a set of worlds, and a world-alternative relation for each agent
\citep{kripke1963semantical}.  An Epik specification defines an alternative relation
on bare events for each agent $a$, which we notate as ${R}_{a}$.  This should be
lifted to a relation $\hat{R}_a$ on worlds.  The basic idea is that when a world $w$
is incremented with an event $e$, in the resulting world $w \diamond e$, epistemic
alternatives for agent $a$ are of the form $w' \diamond e'$, where $w'$ is an
alternative to for $a$ in $w$, and $e'$ is and event-alternative to $e$ for
$a$.\footnote{In this it is important that the event-alternative relation for an
  agent is constant across worlds. We anticipate that the definition given here
  produces results equivalent to what is found in literature on event alternatives in
  dynamic epistemic semantics, though we have not verified this.  That literature
  primarily focuses on mapping an epistemic model for a single time and situation to
  another, and uses general first-order models, rather than guarded string
  models. See \citet{baltag1999logic}, \citet{van2007dynamic}, and articles in
  \citet{ditmarsch2015handbook}.  Previous literature is motivated by epistemic logic
  and AI planning, rather than computable possible worlds models in natural language
  semantics.  } This needs to be implemented in a way that takes account of pre- and
post-conditions for events. For this, our approach is to refer the definition of
well-formed guarded strings.  \ex{raise1} defines a relation on worlds from a
relation on bare events.

\begin{example}  \label{raise1}
  Let $W$ be a set of guarded strings over events P and primitive
  tests B, and ${R}$ be a relation on $P$. The corresponding
  relation $\hat{R}$ on $W$ holds between a guarded string
\(
\alpha_0e_0 ... e_n\alpha_{n+1} 
\)
in $W$ and a guarded string  $q$ iff $q$ is an element of $W$ and is of the form
\(
\alpha_0'e_0' ... e_n'\alpha_{n+1}', 
\), where for $0 \leq n$, $\element{\tuple{e_i,e'_i}}{{R}}$.
\end{example}

@@

\section{The logical language of Epistemic KAT}
The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for KAT, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ is the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  } To this we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  Intuitively, $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1
  ... \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one, $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin ends heads-up.

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}
xs
{\em Standard Epistemic Modalities}
Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$
as in \ex{encodings}.\footnote{Deeper analyisis of the lexical semantics
  of {\em know} requires adding modeling of presupposition
  \citep{collard2018finite}. The grammar fragment in Section 6 does not model
  the presupposition of {\em know}, except (depending on the Epik model) as
  an entailment.}

@@

Different types of reasoners (e.g. accurate, inaccurate,
etc) are modeled using the event-alternatives in an
Epik specification.\footnote{See the discussion of modal axioms \textbf{T} and
  \textbf{D} below.}  The agents in Figure 1 do not always have reliable beliefs,
because of the possibility of secret turning.

\newcommand{\epikspec}{\test,\event,\varphi,\zeta}
\newcommand{\AphiB}{\mathcal A^\varphi_{\mathsf B}}

{\em Guarded String Interpretation.} A term $p$ of the logical language is interpreted as a set of guarded
strings $\sem{p}^{\epikspec}$, where superscript captures dependence
on an Epik specification.  Figure \ref{fig:interp2} defines the
interpretation.  The interpretation $\sem{ 1
}^{\epikspec}$ of the multiplicative identity 1 is the set of atoms that
satisfy the state constraint $\varphi$.  Where $b$ is a primitive
Boolean, $\sem{ b }^{\epikspec}$ is the set of atoms that satisfy the
state constraint and where $b$ is true.  Where $e$ is a primitive
event, $\sem{ e }^{\epikspec}$ is the set of guarded strings that have
the form of $e$ flanked by compatible atoms, as determined by the
event formula $\zeta_e$. The product $pq$ is
interpreted with fusion product raised to sets of guarded strings.
Kleene star is interpreted as the union of exponents ($p^n$ is
the $n$-times product of $p$ with itself, with $p^0 = 1$).
Propositional complement is complement relative to the set of worlds.
The epistemic formula $\Diamond_a p$ is interpreted with Kripke
semantics for epistemic modality, as the pre-image of $p$ 
under the world-alternative relation $\hat{R}_a$.

@@

\label{sec:axioms}

{\em Axiomatic Classification}. To situate our logic as a modal logic, consieer the soundness of
the standard modal axioms given our semantics \citep{hughes1996new}. Some of
these standard axioms (see \ex{valid-axioms}) hold all the time, i.e. they
are \emph{valid}, and the remaining axioms are valid when $\hat{R}_a$
has a certain shape. The axioms in \ex{conditionally-valid-axioms}
have a nontrivial condition validity condition.

@@

The space of worlds is a set of ordinary (as opposed to guarded) strings.  Bit
sequences (sequences of $0$'s and $1$'s) encode atoms, and as before, these alternate
with event symbols to encode a world.  In this construction, $w_3$ of the example is
the string {\tt{10\hspace{2pt}a1\hspace{2pt}10\hspace{2pt}b1\hspace{2pt}10}}.

@@

@@

Translating the Epik terms $0$, $1$, $b$, and $e$ are
straightforward: we simply convert the atoms as previously described,
decorate the events $e$ with their compatible atoms. For example $a_1$ becomes
an Fst term denoting
$\{01a_101\}$. Fst has built-in operations of union (\texttt{|}),
and intersection (\texttt{\&}), which define
the sum and intersection operations in the guarded string algebra.
Fst set difference (\texttt{-}) is used to define propositional complement as the
difference between the set of worlds and the argument.

Defining KAT product using Fst's set-lifted string concatenation (denoted by
juxatposition $\texttt{X} \texttt{Y}$) requires more care. Naively concatenating
strings with atoms (Boolean vectors) at both ends doubles atoms at the juncture, and
does not enforce the requisite atom equality.  To implement KAT product, we define
the binary operation \texttt{Cn}, which concatenates strings in the string algebra,
removes strings with non-matching atoms, and then deletes the second of two atoms to
create a set of well-formed guarded strings.  See Figure \ref{fstfigure1}. {\tt Wf0}
is the set of ordinary strings that does not contain unequal pairs of atoms, as
defined using Fst's containment operator \texttt{\$}. The \texttt{Squash} relation
uses Fst's rewrite notation to delete atoms (elements of \texttt{St}) that are
preceded by another atom. \footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted in that the
  complement and set difference for non-equal length relations is not defined.
  Epistemic alterative relations are equal-length relations.} This relation is
applied via the relational composition (\texttt{.o.}) and codomain (\texttt{.l})
operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

KAT Kleene plus is defined in a similar way using Kleene plus in the string algebra,
with checks for equality of atoms and deletion of atoms. See Figure \ref{fstfigure1}.
KAT Kleene star is definied from KAT Kleene plus and the multiplicative identity,
which is the set of well-formed atoms \texttt{St}.

@@

\begin{hangingpar}
{\tt{define RelKst(R)
    [St\hspace{1pt}.x.\hspace{1pt}St]\hspace{1pt}|\hspace{1pt}Kpl(X);}}\\ The Fst
operation {\tt{.x.}} is Cartesian product.  {\tt{R.i}} is the inverse of relation
{\tt{R}}.
\end{hangingpar}
\end{tcolorbox}
\caption{Definition in Fst of the Kleene concatenation closure of
  a relation between guarded strings.}
\label{relclosure}
\end{figure}

@@

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for
each agent. This determines a relation between bare events decorated with compatible
atoms. 
In Fst, we use the closure of the concatenation
product operation on relations to lift a relation on decorated events for an agent to
the corresponding relation on worlds.  The concatenation product $RS$ of two
relations $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where
$x_1 \mathrel R y_1$, and $x_2 \mathrel S y_2$.  In Fst, {\tt{R+}} is the closure of relation {\tt R} with
respect to this operation. Figure \ref{relclosure} defines the corresponding
operation on sets of guarded strings as encoded in Fst.\footnote{Relation concatenation in Fst
  differs from relation composition (.o.), and the closure under discussion here is
  the closure of the former rather than the latter.}  The epistemic alternative
relation on worlds for an agent is then defined as the KAT relational concatenation
closure {\tt RelKst} of the decorated-event alternative relation for the agent.
